Open Data

### 2020-05-24 draft_chapter.Rmd
another read through some errors tho :(

Access to data in crime and place research, and even criminology more broadly, has traditionally been reserved for those who have the means to collect fresh data themselves, pay for access, or obtain data through formal data sharing agreements. Even when access is granted, the usage of these data often comes with conditions that circumscribe how the data can be used through licensing or policy [@kitchin2014data]. The public dissemination of findings which emerge from analysis might even be subject to restrictions. This can lead to unequal access, controlled usage and curb the diffusion of findings, severely limiting the insight that can be attained from data.

Open data are a response to these shortcomings, broadening access and participation in research, removing the requirement for permissions, formal agreements and negotiations [@manovich2011trending]. Not only this, but open data can also lead to all sorts of novel insight within crime and place research, tapping into constructs and processes which are difficult to capture through surveys, interviews and other traditional measures [@solymosi2018role]. As such, it is important that social scientists, researchers, crime analysts, and others interested in making sense of the social world around them have the skills and know-how to access, interpret, critique and analyse open data sets.

This chapter aims to lay the groundwork for developing such skills by providing a framework to approach and meaningfully interpret and engage with open data. The chapter also offers a practical hands-on guide to accessing, wrangling, and analysing different sources of open data in order to draw conclusions about crime and place. We achieve this by first giving a background on open data, discussing the key types of data out there, critically assessing the strengths and limitations of each, and building a how-to guide checklist for researchers to follow when approaching an open data source. Finally, we work through an example that shows how to access, wrangle, link, and interpret open data from various sources, providing a template which can be applied in other research areas.

@@

- **Availability and Access:** the data must be accessible via a public domain, at no more than a reasonable reproduction cost, and through a convenient medium, such as the internet. Ideally, it should be machine-readable and modifiable. For example, it should be downloadable from a website as an unlocked spreadsheet, or JSON file, rather than presented as a summary table in a PDF document.
- **Re-use and Redistribution:** the data must be provided under conditions that allow free use, such as modification, separation, or compilation of the data, and permit re-use and redistribution including the intermixing (merging) with other datasets.
- **Universal Participation:** everyone should be allowed to use, re-use and redistribute the data and its derivatives without restriction. For instance, restrictions that only allow non-commercial or educational usage would not constitute open data.

In short, 'open data' should be data that are readily and publicly available, in a usable format, and without restrictions on usage, modification and re-distribution. Open data practices, and the extent to which it is a reality, vary both within and between countries. For example, the United States has a history of making public sector data sets openly available. The United Kingdom can be more strict, releasing data under a licence ('Open Government Licence'), with some data requiring a fee [@kitchin2014data]. The _Global Open Data Index_ [@globopendata] is one tool for tracking the annual global benchmark for publication of open government data by country. It provides a rank which indicates how well governments across the world follow open data practices on aggregate, but also according to specific domains. Many of these have direct relevance to crime and place research, such as administrative boundaries and geographic locations (e.g. coordinates).

@@

Now we have a reasonable idea about what open data is in a broad sense, we can consider the different types of open data which you might come across. Here, we formulate a typology of open data sets based on their origin to help provide an overview.

@@

Much of the open data movement has focused on information generated by local and national state agencies, or publicly funded research. Open data has become a key component of transparency and accountability in government, given that much of the data collected has been funded by the public purse [@kitchin2014data]. In some countries, the national statistics body is independent from government, which serves to improve accountability and increase public trust in the data being collated and published. The _Global Open Data Index_ introduced earlier focuses on identifying the gaps in governmental organizations, encouraging them think about how public sector information can become more usable, and ultimately, more impactful [@globopendata].

As researchers studying crime and place, public sector data is particularly important. Many variables thought to explain the non-random distribution of crime, offenders and victims in space are collated and published by governmental bodies. For instance, demographic characteristics about resident populations can be collected via national censuses, with data usually released at aggregations which constitute 'neighborhoods' - a theoretically important scale in criminology. Data describing urban structure, such as street networks, are often generated and distributed by public bodies, and can be used in street syntax analysis to assess whether the configuration of streets dictates things like violent crime victimisation [@streetsyntax2017]. In many countries, representative victimisation surveys are openly available for download. This is not only useful for substantive research (e.g. quantifying long-term crime trends), but also scrutiny over reliability and accuracy in other forms of open data, such as police recorded crime records. Both of these examples represent an interesting challenge to open data principles, since data tends to be anonymised to some degree before being made available to the public. Access to the raw records is dependent on a number of requirements (e.g. background checks, contractual agreements, ICT security measures) which might exclude many individuals due to cost and practicality.

@@

Opening up data generated by the private sector represents a significant challenge, largely due to the proprietary value to its creators [@kitchin2014data]. The aims and objectives of private companies, who have a duty to shareholders and operate in a competitive market environment, differ considerably from local and central government. That said, some datasets are released openly by private sector organizations, albeit often only a subset of what would be available as a paying customer. The website for ArcGIS, a piece of software maintained by the Environmental Systems Research Institute (ESRI), a private company, host a number of open geographic datasets on their online cloud platform (https://hub.arcgis.com/search). Indeed, many papers have made use of data collected or distributed by private organizations to explore crime and place, such as Google Street View images [@langton2017residential] and Twitter [@malleson2015impact]. In fact, you will learn to use the free Twitter API for such purposes in this book (see Chapter 6, Topic 7).

@@

Finally, there is a specific group of open data sources that collate data collectively, generated by large groups of individuals who do not specifically belong to any organization, but instead work together on a collaborative project. These are crowdsourced data sets. Examples include Wikipedia (https://www.wikipedia.org/), an online encyclopedia where anyone can contribute, and Flickr (https://www.flickr.com/) an online photo gallery where people can upload and tag their photos. Often, the organizations who maintain and monitor these data collection activities are charities or non-governmental organizations that do not operate for profit, but instead provide some form of social good. An example is the online reporting platform FixMyStreet (https://www.fixmystreet.com/), where people can report problems such as instances of graffiti, vandalism or environment issues. The data has been used to explore signal crimes theory, for instance, offering insight into people's experiences with incivilities [@solymosi2018crowdsourcing]. Another topic in this book (see Chapter 6, Topic 6) discusses the merits and pitfalls of crowdsourced data, and what to watch out for when analysing open data of this type.

@@

Firstly, one of the biggest obstacles to collecting open data is the ability to interpret what the data means (e.g. applying a framework for its analysis), and computational skills in data scraping, wrangling and cleaning, in order to transform it into a usable format for research [@boyd2012critical]. All sorts of open data remain unanalyzed, and or inaccessible, to large portions of the population, simply because they lack the skills and know-how. Although this is certainly an obstacle for data usage more generally, it represents a key challenge to governmental bodies, in particular, on which there is an onus to ensure transparency and facilitate scrutiny. Private companies, especially those who generate open data as a by-product of their primary activity (e.g. Twitter), have little responsibility or pressure to ensure that their data is accessible and usable to the general public. Moreover, in both public and private spheres, licences and conditions on open data are not necessarily concrete, and might be subject to change with little or no notice. This is an important consideration when planning and running long-term research projects.

Secondly, despite the merits of open data in terms of remedying pitfalls in traditional survey methods, there are other threats to validity that may emerge. For instance, whilst many online resources offer an 'honest' depiction of society, uncaptured by surveys, researchers should be careful in interpreting people's communication online as completely authentic [@manovich2011trending]. This could be a result of individuals willingly managing and 'curating' their online presence [@ellison2006managing], or because of wider issues such as government censorship, or cultural norms around particular topics such as sexual assault. Relatedly, researchers should be aware of issues over sampling, and ultimately, the generalizability of findings that emerge from the analysis of open data. In the case of crowdsourced data, the sample is not randomly drawn from a population, but rather, it is self-selected, giving way for people willing to discuss or contribute to a particular issue, which introduces a degree of bias [@longley2012geodemographics]. Specifically, contributors tend to be men, between the ages of 20-50, with a college or university degree [@budhathoki2010participants; @haklay2010good]. Contributions to resources such as Open Street Map, which we look at in the practical exercise later, are correlated with contextual characteristics such as poverty and population density, and as such, coverage is non-uniformly distributed across urban areas [@mashhadi2013putting]. These are important to keep in mind (and be transparent about) when reporting findings based on analysis of such data.

Open data can be messy, biased and noisey, but criminology (and social sciences more generally) can benefit immeasurably from its use. Only through open data can public sector bodies be held to account, research be transparent and reproducible, and participation in data analysis universal. In crime and place research, both dependent variables (e.g. police recorded crime incidents, victimization rates) and independent variables (e.g. demographic characteristics, ambient population estimates) can be sourced from open data, whether public, private or crowdsourced. With a critical, engaged and considered approach to conducting research with open data, criminology can become a leading force in open and reproducible social science. In sum, researchers and analysts using open data must first ask critical questions:

@@

With this in mind, the following exercise utilizes mutliple sources of open data on public transport and police recorded crime in London, England, to explore crime and place.

@@

In this exercise we will demonstrate a number of different skills, including:

- Accessing open data using direct downloads and Application Programming Interfaces (APIs).
- Cleaning and wrangling data for analysis.
- Visualizing open spatial data.
- Comparing different sources of open data.
- Engaging with the critical questions outline above to better understand open data.

@@

In demonstrating these skills, we will be answering two key research questions. The first is a substantive empirical investigation into crime in and around London Underground stations, in the spirit of existing research into transport crime (see newton2014above). Studies in Sweden, for instance, have shown the importance of environmental and neighborhood characteristics in determining crime concentrations at underground stations in Stockholm, along with the positioning of stations on the line [@ceccato2013security]. Investigations have also taken place into crime concentrations at bus stops in Los Angeles [@loukaitou1999hot] and the impact of intensive policing along bus corridors in Merseyside, England [@newton2004crime], amongst others. As such, we can pose the following research question:

- RQ1: To what extent does crime cluster in and around London Underground stations?

The second is more methodological, and relates to our considered and critical usage of open data, by exploring the validity of different open data sources. As such, we can pose:

- RQ2: How comparable are public sector and crowdsourced open data on London underground stations?

First, we will go through how to access the open data used in the exercises.

@@

The simplest way that open data can be made available is through direct download from a website. This is the case for open police recorded crime data in England and Wales. It can be accessed and downloaded using an online web portal (https://data.police.uk/) under Open Government Licence. Using the "Data" tab, anyone from researchers to members of the public can select the time period and police force of interest, and the type of information required (e.g. crimes, stop and search, outcomes), and then "Generate file" to create a download link. Data is downloaded in  comma-separated values format (.csv file extension). The website also contains detailed information about data quality, downloads and tools under the "About" tab, which is particularly useful when considering the critical questions posed earlier (where, why, who, what, when). For this exercise, we are going to use British Transport Police data, a force which operates on railways and light-rail systems across the country, for the month of January in 2020. Select the time period using the dropdown menus, the force using the tickboxes, and then generate and download the file. We save this file locally to a subfolder named "data", and during the exercise, the data is loaded into R from this folder.

@@

Another way of downloading open data is through an application programming interface (API). This is a tool which defines an interface for a program to interact with a software component or system, for example, what sort of requests or calls can be made to it, and how these can be made. Here, we are using the term 'API' to denote tools created by an open data provider to give access to different subsets of their content. APIs facilitate scripted and programmatic extraction of content, as permitted by the API provider [@olmedilla2016harvesting]. APIs can take many different forms and be of varying quality and usefulness [@foster2016big]. For the purposes of accessing open data from the web, we are specifically talking about _RESTful_ APIs. The 'REST' stands for Representational State Transfer. These APIs work directly over the web, which means users can play with the API with relative ease in order to understand how it works [@foster2016big].

Here, we will use the example of a local government organization responsible for transport, Transport for London (TfL). One of the main responsibilities of TfL is to oversee various rail networks in Greater London, including the London Underground. We will use open data from TfL to practice how to construct HTTP (i.e. web link) requests via their API, and then parse the data which is returned. Much like how downloading data directly requires access to a data dictionary to make sense of the data (e.g. understanding its origins and flaws), we must find a similar document to understand how the API we plan on using works. Details about the TfL API are provided through their open data page (https://tfl.gov.uk/info-for/open-data-users/unified-api), which explains what data is available, and how the API is designed. Usefully, TfL provide a documentation page with examples of how you can use the HTTP links to make calls to the API (https://api-portal.tfl.gov.uk/docs). We will actually make these calls from within the R environment, but it might be useful (and interesting) to examine how basic queries are constructed using this documentation.

@@

Often, developers who work with APIs will share their code, and release them in the form of a package or module. This is called a _wrapper_ because it uses code that 'wraps' around the API to make it a neater, more usable package. Wrappers remove (or at least lower) many of the obstacles to accessing open data noted earlier. The wrapper can take many forms, such as a Python module, or an R package. It could even be a web interface that provides a graphical user interface (GUI) for accessing the API in question.

To demonstrate this, we will be accessing data from Open Street Map, a database of geospatial information built by a community of mappers, enthusiasts and members of the public, who contribute and maintain data about all sorts of environmental features, such as roads, green spaces, restaurants and railway stations, amongst many other things, all over the world. As such, it is a prime example of 'crowdsourced' open data. You can view the information contributed to Open Street Map using their online mapping platform (https://www.openstreetmap.org/). The result of people's contributions is a database of spatial information rich in local knowledge which provides invaluable information about places and their features, without being subject to strict terms on usage. Open Street Map has two types of wrappers available for its API, a web-based GUI called Overpass Turbo (https://overpass-turbo.eu/), and an R package called `osmdata`. We will focus on their usage from within R, as this is a tool used more widely to conduct reproducible research and statistics in criminology.

## Walk-through

So, let's get started. Recall that we are interested in examining the extent to which crime concentrates in and around London Underground stations, but also the contrasts between the open data available via TfL and Open Street Map, as public sector and crowdsourced open data, respectively.

First, ensure that you have the relevant packages installed in R. Although the main packages used in this demonstration are `rjson` for parsing the TfL API results and `osmdata` for querying the Open Street Map API, we use a number of additional packages for data handling and visualization. If you do not have these packages installed, use the `install.packages()` function prior to loading each one with `library()`.

```{r packages}
library(jsonlite) 
library(osmdata)
library(ggplot2)
library(sf)
library(readr)
library(dplyr)
library(tidyr)
library(patchwork)
```

### Direct request to TfL API

Reading through the TfL API documentation mentioned earlier (https://api-portal.tfl.gov.uk/docs), you will get a feel for the kind of data available. To draw conclusions about the degree to which crime occurs in and around London Underground stations, we first need spatial information so we can plot station locations on a map. In the documentation there an 'API area' (which offer some guidance on the types of data available) called _Stops_. We can see that this includes information on bus stops by route (which are numbered e.g. 24), and underground stations by line (which are named e.g. "Bakerloo"). The example for the underground stations is a URL: https://api.tfl.gov.uk/line/bakerloo/stoppoints. This has quite an intuitive structure, since we first want to access the TfL API, look for data by line, select the name of the line we are interested in, and finally pull information about where the train stops. In fact, you can simply copy and paste this URL into your internet browser, and it should look something like Figure 1.

Although this contains the information we requested, it is in a format called _JSON_. This stands for JavaScript Object Notation. JSON is an open standard way of storing and exchanging data, and will most likely be the format in which data are returned from most API calls. As you can see, it is not intuitive to read in its current format. In R, we can use the `fromJSON()` function from the `rjson` package to parse all this information into a data frame (with rows and columns). It will then become more familiar and usable. All we need to do is input the URL from the TfL API into the `txt` argument of the function, with a bit of help from `readLines()` in base R which is used to read text from URLs.

```{r getapi}
api_call <- fromJSON(txt = readLines("https://api.tfl.gov.uk/line/northern/stoppoints"))
```

This gives us an object (`api_call`) which contains all the information returned by the TfL API. JSON is slightly different to traditional data frames with rows and columns, which we are probably quite familiar with, because the data is nested. For instance, `api_call` is classed as a data frame, but upon viewing the object using `View(api_call)` you will notice that there are lists nested within it. Another way of checking this is looping the `class()` function through all the columns in the data frame using `lapply(api_call, class)`. This demonstrates an important challenge faced by researchers when using open data, because dealing with data in this format can be messy and complicated.

>>>>

Access to data in crime and place research has traditionally been reserved for those who have the means to collect fresh data themselves, pay for access, or obtain data through formal data sharing agreements. Even when access is granted, the usage of these data often comes with conditions that circumscribe how the data can be used through licensing or policy [@kitchin2014data]. Even the public dissemination of findings which emerge from analysis might be subject to restrictions. This can lead to unequal access, controlled usage and curb the diffusion of findings, severely limiting the insight that can be attained from data.

Open data initiatives provide a response to these shortcomings, broadening access and participation in research, removing the requirement for permissions, formal agreements, and negotiations [@manovich2011trending]. Open data can lead to all sorts of novel insight within crime and place research, tapping into constructs and processes which are difficult to capture through surveys, interviews and other traditional measures [@solymosi2018role]. As such, it is important that social scientists, researchers, crime analysts, and others interested in making sense of the world around them have the skills and know-how to access, interpret, critique, and analyse open data sets.

This chapter will outline how to develop such skills by providing a framework to approach and meaningfully interpret open data. The chapter also offers a practical hands-on guide to demonstrate how to access, wrangle, and analyse different sources of open data in order to draw conclusions about crime and place.

<!-- We achieve this by first giving a background on open data, discussing the key types of data out there, critically assessing the strengths and limitations of each, and building a how-to guide checklist for researchers to follow when approaching an open data source. Finally, we work through an example that shows how to access, wrangle, link, and interpret open data from various sources, providing a template which can be applied in other research areas. -->

@@

- **Availability and Access:** the data must be accessible via a public domain, at no more than a reasonable reproduction cost, and through a convenient medium, such as the internet. Ideally, it should be machine-readable and modifiable. For example, it should be downloadable from a website as an unlocked spreadsheet, or JSON file, rather than presented as a summary table in a PDF document.
- **Re-use and Redistribution:** the data must be provided under conditions that allow free use, such as modification, separation, or compilation of the data, and permit re-use and redistribution including the intermixing (merging) with other datasets.
- **Universal Participation:** everyone should be allowed to use, re-use and redistribute the data and its derivatives without restriction. For instance, restrictions that only allow non-commercial or educational usage would *not* constitute open data.

>In short, 'open data' should be data that are readily and publicly available, in a usable format, and without restrictions on usage, modification and re-distribution.

Open data practices, and the extent to which it is a reality, vary both within and between countries. For example, the United States has a history of making public sector data sets openly available. The United Kingdom can be more strict, releasing data under a licence ('Open Government Licence'), with some data requiring a fee [@kitchin2014data]. The _Global Open Data Index_ [@globopendata] is one tool for tracking the annual global benchmark for publication of open government data by country. It provides a rank which indicates how well governments across the world follow open data practices across varoius domains. Many of these have direct relevance to crime and place research, such as administrative boundaries and geographic locations (e.g. coordinates).

@@

Now we have a reasonable idea about what open data is in a broad sense, we can consider the different types of open data which you might come across. Here, we formulate a typology of open data sets based on their origin.

@@

The Public Sector refers to organizations owned and operated by local or federal government with the core aim to provide services to the public. Much of the open data movement has focused on opening up information generated by these local and national state agencies [@kitchin2014data]. These can include national statistics, or administrative data. Another set of data included in this group is data collected by publicly funded research, for example victimisation surveys, housing surveys, and others.

The _Global Open Data Index_ introduced earlier focuses on identifying the gaps in governmental organizations, encouraging them think about how public sector information can become more usable, and ultimately, more impactful [@globopendata].

Specific to crime and place research, data sets of interest in this particular domain might include
<!-- As researchers studying crime and place, public sector data is particularly important. Many variables thought to explain the non-random distribution of crime, offenders and victims in space are collated and published by governmental bodies. For instance, demographic characteristics about resident populations can be collected via national censuses, with data usually released at aggregations which constitute 'neighborhoods' - a theoretically important scale in criminology.  -->
data describing urban structure, such as street networks to assess whether the configuration of streets dictates things like violent crime victimisation [@streetsyntax2017], or victimisation surveys mentioned earlier to quantify citizens' perceived safety and security.

<!-- In many countries, representative victimisation surveys are openly available for download. This is not only useful for substantive research (e.g. quantifying long-term crime trends), but also scrutiny over reliability and accuracy in other forms of open data, such as police recorded crime records. Both of these examples represent an interesting challenge to open data principles, since data tends to be anonymised to some degree before being made available to the public. Access to the raw records is dependent on a number of requirements (e.g. background checks, contractual agreements, ICT security measures) which might exclude many individuals due to cost and practicality. -->

@@

Contrary to public sector, private sector refers to the rest of the economy whihc are not under direct state control. Opening up data generated by the private sector represents a significant challenge, largely due to the proprietary value to its creators [@kitchin2014data]. The aims and objectives of private companies, who have a duty to shareholders and operate in a competitive market environment, differ considerably from local and central government.

That said, some datasets are released openly by private sector organizations, albeit often only a subset of what would be available as a paying customer. The website for ArcGIS, a piece of software maintained by the Environmental Systems Research Institute (ESRI), a private company, host a number of open geographic datasets on their online cloud platform (https://hub.arcgis.com/search). Indeed, many papers have made use of data collected or distributed by private organizations to explore crime and place, such as Google Street View images [@langton2017residential] and Twitter [@malleson2015impact]. In fact, you will learn to use the free Twitter API for such purposes in this book (see Chapter 6, Topic 7).

@@

Finally, there is a specific group of open data sources that collate data collectively, generated by large groups of individuals who do not specifically belong to any organization, but instead work together on a collaborative project. These are crowdsourced data sets. Examples include Wikipedia (https://www.wikipedia.org/), an online encyclopedia where anyone can contribute, or Flickr (https://www.flickr.com/) an online photo gallery where people can upload and tag their photos. Often, the organizations who maintain and monitor these data collection activities are charities or non-governmental organizations that do not operate for profit, but instead provide some form of social good. An example is the online reporting platform FixMyStreet (https://www.fixmystreet.com/), where people can report problems such as instances of graffiti, vandalism or environment issues. The data has been used to explore signal crimes theory, for instance, offering insight into people's experiences with incivilities [@solymosi2018crowdsourcing]. Another topic in this book (see Chapter 6, Topic 6) discusses the merits and pitfalls of crowdsourced data, and what to watch out for when analysing open data of this type.

@@

Firstly, one of the biggest obstacles to collecting open data is the ability to interpret what the data means (e.g. applying a framework for its analysis), and computational skills in data scraping, wrangling and cleaning, in order to transform it into a usable format for research [@boyd2012critical]. All sorts of open data remain inaccessible to people who may lack the skills and know-how to acquire them. Although this is certainly an obstacle for data usage more generally, it represents a key challenge to governmental bodies, in particular, on which there is an onus to ensure transparency and facilitate scrutiny. Private companies, especially those who generate open data as a by-product of their primary activity (e.g. Twitter), have little responsibility or pressure to ensure that their data is accessible and usable to the general public. Moreover, in both public and private spheres, licences and conditions on open data are not necessarily concrete, and might be subject to change with little or no notice. This is an important consideration when planning and running long-term research projects which involve open data.

Secondly, despite the merits of open data in terms of remedying pitfalls in traditional survey methods, there are other threats to validity that may emerge. For instance, whilst many online resources offer an 'honest' depiction of society, uncaptured by surveys, researchers should be careful in interpreting people's communication online as completely authentic [@manovich2011trending]. This could be a result of individuals willingly managing and 'curating' their online presence [@ellison2006managing], or because of wider issues such as government censorship, or cultural norms around particular topics, particularly sensitive topics of interest to researchers of crime, such as sexual assault or drug use.

Relatedly, researchers should be aware of issues over sampling, and ultimately, the generalizability of findings that emerge from the analysis of open data. In the case of crowdsourced data, the sample is not randomly drawn from a population, but rather, it is self-selected, giving way for people willing to discuss or contribute to a particular issue, which introduces a degree of bias [@longley2012geodemographics]. Specifically, contributors tend to be men, between the ages of 20-50, with a college or university degree [@budhathoki2010participants; @haklay2010good]. Contributions to resources such as Open Street Map (which we look at in the practical exercise later) are correlated with contextual characteristics such as poverty and population density, and as such, coverage is non-uniformly distributed across urban areas [@mashhadi2013putting]. These are important to keep in mind (and be transparent about) when reporting findings based on analysis of such data.

### What can be done?

Open data can be messy, biased and noisey, but criminology (and social sciences more generally) can benefit immeasurably from its use. Only through open data can public sector bodies be held to account, research be transparent and reproducible, and participation in data analysis universal. In crime and place research, both dependent variables (e.g. police recorded crime incidents, victimization rates) and independent variables (e.g. demographic characteristics, ambient population estimates) can be sourced from open data, whether public, private or crowdsourced.

So what can we do to make sure we make good use of these data? With a critical, engaged and considered approach to conducting research with open data, criminology can become a leading force in open and reproducible social science. In sum, researchers and analysts using open data must first ask critical questions:

@@

The answers to these questions will put the data sources in a context of understanding, and ensure that the researchers use them appropriately and with care.

With this in mind, we now move on to a practical exercise, where we will utilize mutliple sources of open data, and explore police recorded crime on and around public transport in London, England.

@@

In this exercise you will acquire a number of different skills, including:

- Accessing open data using 3 different methods: 
    + direct downloads 
    + direct calls to an Application Programming Interface (API)
    + using an API wrapper.
- Cleaning, wrangling, and visualizing open spatial data.
- Comparing different sources of open data.
- How to engage with the critical 'Where', 'Why', 'Who', 'What', 'When' questions to better understand open data.

@@

In this exercise we will explore crime in and around London Underground stations. We know that there are important environmental features when it comes to public transport areas being more or less criminogenic environments. Studies in Sweden, for instance, have shown the importance of environmental and neighborhood characteristics in determining crime concentrations at underground stations in Stockholm, along with the positioning of stations on the line [@ceccato2013security]. Similar work has been done exploring bus stops in Los Angeles [@loukaitou1999hot] and the impact of intensive policing along bus corridors in Merseyside, England [@newton2004crime], amongst others.

Now we will consider the case of London. Specifically, we will examine to what extent does crime cluster in and around London Underground stations?

We will use various sources of open data to answer this question, and this will allow us to explore the strengths and limitations across public sector and crowdsourced open data sources.

@@

The simplest way that open data can be made available is through direct download from a website. In such a case, you can visit a website, select some parameters, and save a file containing the data you requested locally on your computer.

In the United Kingdom, police recorded crime data in England and Wales can be accessed this way using an online web portal (https://data.police.uk/) under Open Government Licence.

Visit the website above, and you will see a welcome message, and 6 tabs across the top of it, which should read "Home", "Data", "API", "Changelog", "Contact", and "About".

Before we download any data, we can learn more about it by clicking on the "About" tab. This brings up a lot of information that it is important to carefully read through to be able to answer the where, why, who, what, when questions posed earlier. Take a moment to read through, and try to take notes on what you think might be relevant for your analysis. For example, if we want to map crimes, we might want to explore if there is any type of "anonymisation" that might take place before the data are released (to protect the privcy of the victims for example). If you read the about page here, you might find the following note:

> Location Anonymisation
The latitude and longitude locations of Crime and ASB incidents published on this site always represent the approximate location of a crime — not the exact place that it happened.

This indicates that although we get a latitude and longitude coordinate with each crime event, it may only be approximate, and this may have implications for our findings later!

To then actually download some data, move on to clicking on the "Data" tab. This should open for you a page titled "Data downloads" where you can see another 5 tabs: "Custom download", "Archive", "Boundaries", "Open data", and "Statistical data".  Staying on this "Custom download" page you can make some selections on what sort of data you wanted to download.

We can select the time period and police force of interest, and the type of information required (e.g. crimes, stop and search, outcomes). We are also informed that the data are downloaded in comma-separated values format (.csv file extension), which is in line with machine-readable, easy to manipulate data format requirements.

For this exercise, we are going to use British Transport Police data, a force which operates on railways and light-rail systems across the country, for the month of January in 2020. Select the time period using the dropdown menus, the force using the tickboxes, and then generate and download the file.

Save this file locally (in your working directory) to a subfolder named "data", and then you can load it in with the `read_csv()` command from the `readr` package (remember to load it with `library(readr)`).

```{r crimeload1}
library(readr)
btp_df <- read_csv("data/2020-01-btp-street.csv")
```

Note: If you didn't manage to follow along with the download instructions, you can also access this dataset from our github page:

```{r crimeload_gh, eval = FALSE}
btp_df <- read_csv("https://raw.githubusercontent.com/langtonhugh/osm_crim/master/data/2019-01-greater-manchester-street.csv")
```

You should have now a dataframe with `r nrow(btp_df)` crimes in them ready for you to explore!

@@

Another way of downloading open data is through an application programming interface (API). This is a tool which defines an interface for a program to interact with a software component, for example, what sort of requests or calls can be made to it, and how these can be made. Here, we are using the term 'API' to denote tools created by an open data provider to give access to different subsets of their content. Such APIs facilitate scripted and programmatic extraction of content, as permitted by the API provider [@olmedilla2016harvesting]. APIs can take many different forms and be of varying quality and usefulness [@foster2016big]. For the purposes of accessing open data from the web, we are specifically talking about _RESTful_ APIs. The 'REST' stands for Representational State Transfer. These APIs work directly over the web, which means users can play with the API with relative ease in order to understand how it works [@foster2016big].

Here we will make a direct request to the API created by London's local government organization responsible for transport: Transport for London (TfL). TfL oversee the London Underground, and provide access to their open data through their unified API. Much like how we found the "About" page when downloading police data directly, to answer our 'Where', 'Why', 'Who', 'What', 'When' questions, we must find a similar document here. Details about the TfL API are provided through their open data page (https://tfl.gov.uk/info-for/open-data-users/unified-api), which explains what data is available, and how the API is designed.

They further provide a documentation page with examples of how you can use HTTP (i.e. web link) requests to make calls to the API (see: https://api-portal.tfl.gov.uk/docs). If you follow this link you should see a page titled "Our Unified API" and some examples of calls.

To draw conclusions about crimes at London Underground stations, we will need some data specifically about these underground stations.

In the documentation there 'API area' (which offer some guidance on the types of data available). Scroll down and find the example calle for stops _Stops_. In the description you will  see that this call returns information on stops for busses or London Underground lines, and there are two examples (bus route 24 and the Bakerloo line). Take their demonstration url you see for the bakerloo line (https://api.tfl.gov.uk/line/bakerloo/stoppoints) and paste it into your web browser. When you visit this page you should see something like Figure 1:

Although this contains the information we requested, it is not very legible for our human eyes. It in a format called _JSON_. This stands for JavaScript Object Notation. JSON is an open standard way of storing and exchanging data, and will most likely be the format in which data are returned from most API calls. As you can see, it is not intuitive to read in its current format but it is very machine readable friendly, once again in line with this requirement for open data.

Now you won't usually make these calls by pasting them into a web browser, but we wanted for you to see how it works. In readlity, we will actually make these calls from within the R environment, but it might be useful (and interesting) to examine how basic queries are constructed using this documentation. For example, we now know that to request the stops from the "Bakerloo" line, we need the url:

`"https://api.tfl.gov.uk/line/bakerloo/stoppoints"`

Let's say instead we want data for the Northern Line. What do you think that URL will look like ?

Indeed, all you need to do is replace "bakerloo" with "northern", and you got the stops for this route instead. Now let's see how we would make this call in R.

In R, we can use the `fromJSON()` function from the `rjson` package to parse all this information into a data frame (with rows and columns). It will then become more familiar and usable. All we need to do is input the URL from the TfL API into the function, with a bit of help from `readLines()`, another function in base R which is used to read text from URLs.

```{r getapi}
library(rjson)

api_call <- fromJSON(readLines("https://api.tfl.gov.uk/line/northern/stoppoints"))
```

This gives us an object (`api_call`) which contains all the information returned by the TfL API. JSON is slightly different to traditional data frames with rows and columns, which we are probably more familiar with, because the data is nested. For instance, `api_call` is classed as a data frame, but just try  viewing the object using `View(api_call)`.

You will notice that you see a list of lists, which is what we call nested lists. Another way of checking what sort of object this `api_call` object is is by looping the `class()` function through all the columns in the data frame using `lapply(api_call, class)`.

If you run this, you will see that each element in `api_call` is a "list".

This demonstrates an important challenge faced by researchers when using open data, because dealing with data in this format can be messy and complicated, not always a neatly formatted dataframe like the csv from the police.uk open data portal. But we will learn how to manage this data into a more familiar format later.

@@

Finally, the last way we will look at how open data is often shared is by using API wrappers. Often, developers who work with APIs will share their code, and release them in the form of a package or module, so that other people can use it too. This is called a _wrapper_ because it uses code that 'wraps' around the API to make it a neater, more usable package. Wrappers remove (or at least lower) many of the obstacles to accessing open data noted earlier. The wrapper can take many forms, such as a Python module, or an R package. It could even be a web interface that provides a graphical user interface (GUI) for accessing the API in question.

To demonstrate this, we will be accessing data from Open Street Map, a database of geospatial information built by a community of mappers, enthusiasts and members of the public, who contribute and maintain data about all sorts of environmental features, such as roads, green spaces, restaurants and railway stations, amongst many other things, all over the world. As such, it is a prime example of 'crowdsourced' open data. You can view the information contributed to Open Street Map using their online mapping platform (https://www.openstreetmap.org/). The result of people's contributions is a database of spatial information rich in local knowledge which provides invaluable information about places and their features, without being subject to strict terms on usage.

Open Street Map has two types of wrappers available for its API, a web-based GUI called Overpass Turbo (https://overpass-turbo.eu/), and an R package called `osmdata`.

If we load and use the package `osmdata` we can use its funcitons to query the Open Street Map API, rather than the API query being made directly (like we did for TfL, using the URL). Once again, we want to have some documentation to be able to refer to about our data and the ways to query it. When it comes to OSM, you can read all about it on their community page: [https://www.openstreetmap.org/about](https://www.openstreetmap.org/about). Regarding the wrapper, as it is an R package, it will come with a handy vignette, which in this case we can find at [https://cran.r-project.org/web/packages/osmdata/vignettes/osmdata.html](https://cran.r-project.org/web/packages/osmdata/vignettes/osmdata.html).

Unlike the TfL API, `osmdata` is an international database, and has lots of data that we might not necessarily need. The first thing we need to specify therefore, is where we want to get data from. To do this, we use the `getbb()` function from the `osmdata` package, which stands for "get bounding box". You can think of the bounding box as a box drawn around an area that we are interested in (in this case, London, UK) which tells the OSM API that we want everything *inside* the box, but nothing *outside* the box.

So how can we name a bounding box specification to define the study region? This can be obtained manually, which requires some existing knowledge about an area using the latitude and longitude coordinates, or you can use a search term. Here, we want to select Greater London, in the UK so we can use the search term "greater london united kingdom". Besides specifying where we want data from, we can also tell the `getbb()` function what format we want the data in. Let's say in this case we want a spatial object, specifically an `sf` polygon object which we name `bb_sf`:

```{r bb}
library(osmdata)

bb_sf <- getbb(place_name = "greater london united kingdom", format_out = "sf_polygon")
```

We now have our study region defined as the administrative boundaries of Greater London, which we can visualize quickly to check it has worked as expected. To visualise we can use the `sf` and `ggplot2` packages.

```{r bbmap}
library(ggplot2)
library(sf)

ggplot(data = bb_sf) +
  geom_sf()
```

If you are familiar with the shape of London then this image should be familiar as well! Pleased with our bounding box, we can now move on to scrape data from the Open Street Map API using the `opq()` function. The function name is short for 'Overpass query', which is how users can query the Open Street Map API using search criteria.

Besides specifying what area we want to query with our bounding box object (`bb_sf`) in the `opq()` function, we must also define the feature which we want to pull from the API. Features in Open Street Map are defined through 'keys' and 'values'. Keys are used to describe a broad category of features (e.g. highway, amenity), and values are more specific descriptions (e.g. cycleway, bar). These are tags which contributors to Open Street Map have defined. A useful way to explore these is by using the comprehensive Open Street Map Wiki page on map features (https://wiki.openstreetmap.org/wiki/Map_Features).

We do this using the `add_osm_feature()` function, specifying our key as 'public transport' and our value as 'station'.

We also want to specify what sort of object (what class) to get our data into, and as we're still working with spatial data we stick to the `sf` format, for which the function is `osmdata_sf()`.

And finally, we trim everything down into our bounding box parameters with the `trim_osmdata()` function. We add all these together using a pipe (`%>%`) and make the call like so:
